\documentclass{article}

% Use numeric citations with natbib.
\PassOptionsToPackage{numbers, compress}{natbib}

% NeurIPS 2026 template.
% For submission, keep the default option below. For camera-ready, use e.g.
% \usepackage[main, final]{neurips_2026}
% For arXiv/preprint, use \usepackage[preprint]{neurips_2026}
\usepackage[preprint]{neurips_2026}

% Standard NeurIPS/sample-template packages
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{hyperref}
\usepackage{url}
\usepackage{booktabs}
\usepackage{amsfonts}
\usepackage{nicefrac}
\usepackage{microtype}
\usepackage[table]{xcolor}
\usepackage{wrapfig}
\usepackage[most]{tcolorbox}



% Packages from the original main.tex
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{mathtools}
\usepackage{amsthm}
\usepackage{enumitem}
\usepackage{tabularx}
\usepackage[capitalize,noabbrev]{cleveref}
\usepackage{bm}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{multirow}
\usepackage{dsfont}
\usepackage{soul}
\usepackage[textsize=tiny]{todonotes}
\definecolor{bestgreen}{RGB}{220,245,220}
\definecolor{lightred}{RGB}{255,225,225}
\definecolor{lightyellow}{RGB}{255,245,200}
\definecolor{lightblue}{RGB}{225,240,255}


\newtcolorbox{highlightbox}{
  enhanced,
  breakable,
  colback=lightblue,
  colframe=blue!0!black,
  boxrule=0.8pt,
  arc=7pt,
  left=7pt,
  right=7pt,
  top=7pt,
  bottom=7pt,
  before skip=10pt,
  after skip=10pt
}

\allowdisplaybreaks

\newtcolorbox{chatbox}{
    colback=gray!5!white,
    colframe=gray!75!black,
    arc=5pt,
    boxrule=0.5pt,
    left=5pt,
    right=5pt,
    top=5pt,
    bottom=5pt,
    fontupper=\ttfamily\small
}

% Attempt to make hyperref and algorithmic work together better.
% \newcommand{\theHalgorithm}{\arabic{algorithm}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% THEOREMS
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\theoremstyle{plain}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{assumption}[theorem]{Assumption}
\theoremstyle{remark}
\newtheorem{remark}[theorem]{Remark}


% common notations
\newcommand{\one}{\bm 1}
\newcommand{\zero}{\textbf{0}}
\newcommand{\e}{{\bm e}}
\newcommand{\Var}{\mathsf{Var}}
\newcommand{\LL}{\mathcal L}
\newcommand{\Cov}{\mathrm{Cov}}
\newcommand{\U}{\mathrm{Unit}}
\newcommand{\bdiag}{\mathrm{bdiag}}
\newcommand{\EE}{\mathbb{E}}
\newcommand{\PP}{\mathbb{P}}
\newcommand{\QQ}{\mathbb{Q}}
\newcommand{\nm}[1]{\left\|{#1}\right\|}
\newcommand{\ra}{\rightarrow}
\newcommand{\R}{\mathbb{R}}
\newcommand{\ip}[1]{\left<{#1}\right>}
\newcommand{\eqbr}{\phantom{{}=1}}
\newcommand{\avg}[1]{\frac{1}{#1}\sum_{i=1}^{#1}}

% transformers
\newcommand{\WQK}[1]{\bm W_\mathrm{QK}^{({#1})}}
\newcommand{\WOV}[1]{\bm W_\mathrm{OV}^{({#1})}}
\newcommand{\Att}[2]{\sigma\left(\frac{{#1}{#2}{#1}^\top}{\sqrt{d}}\right)}
\newcommand{\fd}[2]{\frac{\partial{#1}}{\partial{#2}}}
\newcommand{\sd}[2]{\frac{\partial^2{#1}}{\partial{#2}^2}}
\newcommand{\vecr}{\mathrm{vec}_\mathrm{r}}
\newcommand{\vecc}{\mathrm{vec}_\mathrm{c}}
\newcommand{\mask}{\texttt{M}}
\newcommand{\alg}{VLP}

\algnewcommand{\Input}{\item[\textbf{Input:}]}
\algnewcommand{\Output}{\item[\textbf{Output:}]}

\DeclareMathOperator*{\argmin}{argmin}
\DeclareMathOperator*{\argmax}{argmax}
\DeclareMathOperator{\prox}{prox}
\DeclareMathOperator{\proj}{proj}
\DeclareMathOperator{\dom}{dom}
\DeclareMathOperator{\tr}{tr}
\DeclareMathOperator{\Det}{Det}
\DeclareMathOperator{\diag}{diag}
\DeclareMathOperator{\Diag}{Diag}
\DeclareMathOperator{\Rank}{Rank}
\DeclareMathOperator{\spn}{span}
\DeclareMathOperator{\sgn}{sgn}
\DeclareMathOperator{\ind}{\mathds{1}}

\def\vx{{\bm{x}}}
\def\vy{{\bm{y}}}
\def\vs{{\bm{s}}}

% Author comments
\newcommand{\jk}[1]{\textcolor{red}{Jason: #1}}
\newcommand{\sijin}[1]{\textcolor{blue}{[Sijin: #1]}}
\newcommand{\cy}[1]{\textcolor{green}{Chang: #1}}
\newcommand{\jiaxin}[1]{\textcolor{blue}{[Jiaxin: #1]}}
\newcommand{\strblk}[1]{\sethlcolor{black!12}\hl{\texttt{\footnotesize #1}}}

\title{On Confidence-Based Decoding Strategies for Masked Diffusion and Beyond: A Likelihood Optimization Perspective}

% In the default NeurIPS submission mode, the style anonymizes author information.
% For the camera-ready version, compile with the final option above.
\author{%
  Sijin Chen \\
  Princeton University \\
  \texttt{chensj@princeton.edu}
  \And
  Chang Yu \\
  Princeton University \\
  \texttt{cy7438@princeton.edu}
  \And
  Jiaxin Shi \\
  Meta SuperIntelligence Labs
  \And
  Jason M. Klusowski \\
  Princeton University \\
  \texttt{jason.klusowski@princeton.edu}
}

\begin{document}

\maketitle

\begin{abstract}
Masked diffusion language models enable a more flexible generation order than their autoregressive counterparts, but raise the question of what unmasking order is optimal during sampling. Empirically, confidence-based unmasking heuristics perform strongly in practice, but the theoretical basis for their advantages remains unclear. In this work, we develop a theoretical account of confidence-based methods by studying a likelihood optimization problem that searches for the best unmasking policy to maximize sample likelihood. Under planning-depth reduction and distribution approximation, we show that confidence unmasking emerges as the (near-)optimal policy in our framework, justifying its empirical superiority. Building on the likelihood principle, we further propose a trainable policy that learns to optimize the global objective without the previous simplifications. Across experiments on multiple MDMs of different scales spanning text, code, and math generation benchmarks, the trained policy matches or exceeds the performance of confidence-based methods. Overall, our work demonstrates the potential of a likelihood optimization framework for analyzing and developing efficient sampling methods for diffusion language models.
\end{abstract}

\input{text/introduction}
\input{text/preliminaries}
\input{text/theory}
\input{text/method}
\input{text/experiments}

\section{Conclusion}
In this work, we study inference in masked diffusion language models through a likelihood-based formulation that treats the unmasking order as a policy to be optimized. Under local planning and distribution-approximation reductions, this framework explains confidence-based unmasking as an near-optimal policy, with numerical evidence supporting the likelihood-optimality view. Building on the same principle, we introduced Variational Likelihood Policy (\alg{}), a trainable unmasking policy optimized by a variational lower bound, and showed that it matches or improves over uniform and confidence-based unmasking across text, code, and math generation tasks and multiple MDMs at different scales.

This work still has several limitations. The theoretical derivation relies on local reductions and approximation assumptions, while the design of VLP is prototypical and can face inefficiency and instability in training. Future work may improve the optimization of VLP, explore joint training with the underlying diffusion model, and extend the likelihood principle to other discrete diffusion models beyond MDMs.

% \section*{Impact Statement}

% This paper presents work whose goal is to advance the field of machine learning. There are many potential societal consequences of our work, none of which we feel must be specifically highlighted here.

% For NeurIPS final versions, acknowledgments should use the ack environment.
% Do not include acknowledgments in the anonymized submission.
% \begin{ack}
% Acknowledgments go here.
% \end{ack}

\newpage

{\small
\bibliographystyle{plainnat}
\bibliography{main}
}

% \newpage
% \input{checklist.tex}

\newpage
\appendix
\input{text/appendix}


\end{document}
